Metacircular Virtualization \& \\ Practical Nock Interpretation

### 2025-07-07 mss.tex
Fix brackets and noun.
Post section

Hearkening back to \citet{Amin2017}'s collapsed towers of interpreters, the affordances of \lstinline[style=inlinecode]{++mock} as a metacircular interpreter permit us to build a tower of interpreters on top of it without incurring excessive performance overhead.  As the author of \texttt{u3.md} concluded, ``we [even] simply treat Nock proper as a special case of \lstinline[style=inlinecode]{mock}.''

What would such an interpreter look like, and how would it relate to actual operation on a runtime in practice?  Let us extend \lstinline[style=inlinecode]{++mink} to implement a logical loobean operator.  (Call the modified gate \lstinline[style=inlinecode]{++pink}.)  This operation will be invoked with a new fake opcode 13.  Under the hood, opcode 13 will simply supply logically negate a value (via an opcode 6), much as how opcode 12 supplies a consistent Nock formula for the scry operation.  (Similar to \lstinline[style=inlinecode]{[0 0]}, we will be able to use \lstinline[style=inlinecode]{[13 1 2]} to initiate a crash if desired.)

@@

This Nock formula produces a working virtualized Nock interpreter derived from Hoon code with some hand-tuned design.  The input function signature is \lstinline[style=inlinecode]{[subject=* formula=*]} with output of the form \lstinline[style=inlinecode]{[@ product=*]}; if the head is \lstinline[style=inlinecode]{%0}, then the product is a valid Nock noun, while a head of \lstinline[style=inlinecode]{%bail} indicates a crash in the evaluation.  Opcode 9 calls may be addressed into the core as subject, and the formula itself is complete with an empty subject.

@@

[1
 :: +mul multiplication of two atoms
 [8 [1 1 1]

@@

 [:: +add addition of two atoms
  [8 [1 0 0]

@@

    [:: +list type definition
      [8

@@

                [[6 [5 [1 0] 0 12]

@@

@@

          [1 [:: +frag fragmentary subtree of noun
              8 [8 [9 342 0 127] 9 2 10 [6 0 60] 0 2]
                [14 [8 [9 87 0 255]
                 0 3] 9 2 0 1] 0 1]

@@

              8 [9 2 10
                 [26

@@

                 10 [12 8 [9 87 0 127]
                    9 2 10 [6 0 60] 0 2] 0 3]

@@

                6 [5 [1 2] 0 6]
                  [[1 0] [0 5] 0 245]
                    [[1 0] 0 244]

@@

              0 5] 0 1]

@@

          [6 [3 0 29] [3 0 58] 1 1]
          [8 [9 11 10 [29 0 58] 0 1]

@@

              [8 [9 11 10 [29 0 123] 0 3]

@@

          [6 [3 0 29]
             [6 [5 [1 0] 0 58]
                [6 [3 0 59] [1 1] 1 0]

@@

          [8 [8 [9 4 0 1]
                9 2 10 [6 [0 123] 0 60] 0 2]

@@

          [6 [3 0 29]
             [5 [1 1] 0 58] 1 1]
          [[1 0] 0 59]

@@

          [6 [3 0 29]
             [6 [5 [1 2] 0 58] [3 0 59] 1 1] 1 1]

@@

             [8
              [9 11 10 [29 0 247] 0 3]
              6
              [5 [1 0] 0 4]
              [9 11 10 [14 [0 13] 0 5] 0 7]

@@

          [6 [3 0 29]
             [5 [1 3] 0 58] 1 1]
          [8 [9 11 10 [29 0 59] 0 1]

@@

          [6 [3 0 29]
             [5 [1 4] 0 58] 1 1]
          [8
            [9 11 10 [29 0 59] 0 1]
            6
            [5 [1 0] 0 4]

@@

          [6 [3 0 29]
             [6 [5 [1 5] 0 58] [3 0 59] 1 1] 1 1]
          [ 8
            [9 11 10 [29 0 118] 0 1]
            6
            [5 [1 0] 0 4]
            [ 8
              [9 11 10 [29 0 247] 0 3]
              6
              [5 [1 0] 0 4]

@@

          [6 [3 0 29]
             [6 [5 [1 6] 0 58]
                [6 [3 0 59] [3 0 119] 1 1]
                1 1] 1 1]

@@

             6
             [5 [1 0] 0 4]

@@

                [9 11 10 [29 0 494] 0 3]

@@

                  [9 11 10 [29 0 495] 0 3]

@@

::              11
::              [1.936.945.012 1 0]
::              1 1.818.845.538 0

@@

          [6 [3 0 29]
             [6 [5 [1 7] 0 58] [3 0 59] 1 1] 1 1]

@@

               [9 11 10 [14 [0 5] 0 247] 0 3]

@@

          [6 [3 0 29]
             [6 [5 [1 8] 0 58] [3 0 59] 1 1] 1 1]

@@

               [9 11 10 [14 [[0 5] 0 60] 0 247] 0 3]

@@

          [6 [3 0 29]
             [6 [5 [1 9] 0 58]
                [6 [3 0 59]
                   [6 [3 0 118] [1 1] 1 0]

@@

          [8 [9 11 10 [29 0 119] 0 1]

@@

               [8 [8 [9 4 0 3]
                     9 2 10 [6 [0 502] 0 13] 0 2]

@@

              9 11 10 [14 [0 13] 0 5]

@@

          [6 [3 0 29]
             [6 [5 [1 10] 0 58]
                [6 [3 0 59]
                   [6 [3 0 118]
                      [6 [3 0 236] [1 1] 1 0]

@@

          [6 [5 [1 0] 0 236]

@@

             8 [9 11 10 [29 0 119] 0 1]

@@

               [8 [9 11 10 [29 0 493] 0 3]

@@

               [8 [8 [9 10 0 7]
                     9 2 10 [6 [0 2.028] [0 29] 0 13]

@@

          [6 [3 0 29]
             [6 [5 [1 11] 0 58]
                [6 [3 0 59]
                   [6 [3 0 118]

@@

                2 [0 60] [1 11] [0 246] [1 1] 0 5]

@@

::          11
::          [1.936.945.012 1 0]

@@

        9 11 0 1]

@@

@@

                  8 [7 [0 7] 8 [9 22 0 7]

@@

TODO there's a missing open bracket I need to chase down

>>>>

\section{Nock in Nock}

Hearkening back to \citet{Amin2017}'s collapsed towers of interpreters, the affordances of \lstinline[style=inlinecode]{++mock} as a metacircular interpreter permit us to build a tower of interpreters on top of it without incurring excessive performance overhead.  As the author of \texttt{u3.md} concluded, ``we [even] simply treat Nock proper as a special case of \lstinline[style=inlinecode]{mock}.''  The conventional \lstinline[style=inlinecode]{++mink} gate is built in Hoon, but what would happen if we stripped it down and built it in Nock itself with the minimal necessary machinery?  What would such an interpreter look like, and how would it relate to actual operation on a runtime in practice?

Let us extend \lstinline[style=inlinecode]{++mink} to implement a logical loobean operator.  (Call the modified gate \lstinline[style=inlinecode]{++pink}.)  This operation will be invoked with a new fake opcode 13.  Under the hood, opcode 13 will simply supply logically negate a value (via an opcode 6), much as how opcode 12 supplies a consistent Nock formula for the scry operation.  (Similar to \lstinline[style=inlinecode]{[0 0]}, we will be able to use \lstinline[style=inlinecode]{[13 1 2]} to initiate a crash if desired.)

@@

This Nock formula produces a working virtualized Nock interpreter derived from Hoon code with some hand-tuned design.  The input function signature is \lstinline[style=inlinecode]{[subject=* formula=*]} with output of the form \lstinline[style=inlinecode]{[@ product=*]}; if the head is \lstinline[style=inlinecode]{%0}, then the product is a valid Nock noun, while a head of \lstinline[style=inlinecode]{%bail} indicates a crash in the evaluation.  Opcode 9 calls may be addressed into the core as subject, and the formula itself is complete with an empty subject.  (Note as well that the \lstinline[style=inlinecode]{%bail} return may itself be the natural result of a valid Nock computation, which would be avoidable if we switched to a unit return instead.)

@@

[[1
  [:: +mul multiplication of two atoms

@@

  [[:: +add addition of two atoms

@@

     [[8
       :: +list type definition

@@

                [[6 [5 [1 0] 0 12] [1 0] 0 0]

@@

              6 [5 [1 0] 0 6] [1 0] 0 0]

@@

@@

               8 [9 2 10 [26

@@

                            10 [12 8 [9 87 0 63]

@@

                   8 [8 [9 342 0 15]
                        9 2 10 [6 0 60] 0 2]
                       [7 [0 3] [1 0] [0 5] 0 117]
                         [7 [0 3] [1 0] [0 116] 0 5]

@@

@@

           [6 [3 0 13] [3 0 26] 1 1]
           [8 [9 2 10 [13 0 26] 0 1]

@@

               [8 [9 2 10 [13 0 59] 0 3]

@@

           [6 [3 0 13]
              [6 [5 [1 0] 0 26]
                 [6 [3 0 27] [1 1] 1 0]

@@

           [8 [8 [9 22 0 7]
                 9 2 10 [6 [0 59] 0 28] 0 2]

@@

           [6 [3 0 13]
              [5 [1 1] 0 26] 1 1]

@@

           [6 [3 0 13]
              [6 [5 [1 2] 0 26] [3 0 27] 1 1] 1 1]
           [8 [9 2 10 [13 0 54] 0 1]

@@

                [8 [9 2 10 [13 0 119] 0 3]
                     [9 2 10 [6 [0 13] 0 5] 0 7]

@@

           [6 [3 0 13]
              [5 [1 3] 0 26] 1 1]
           [8 [9 2 10 [13 0 27] 0 1]

@@

           [6 [3 0 13]
              [5 [1 4] 0 26]
           [8 [9 2 10 [13 0 27] 0 1]

@@

           [6 [3 0 13]
              [6 [5 [1 5] 0 26] [3 0 27] 1 1] 1 1]
                [8 [9 2 10 [13 0 119] 0 3]

@@

           [6 [3 0 13]
              [6 [5 [1 6] 0 26]
                 [6 [3 0 27] [3 0 55] 1 1] 1 1] 1 1]

@@

@@

                   [9 2 10 [13 0 238] 0 3]

@@

                     [9 2 10 [13 0 239] 0 3]

@@

@@

           [6 [3 0 13]
              [6 [5 [1 7] 0 26] [3 0 27] 1 1] 1 1]
           [8 [9 2 10 [13 0 54] 0 1]

@@

                [9 2 10 [6 [0 5] 0 119] 0 3]

@@

           [6 [3 0 13]
              [6 [5 [1 8] 0 26] [3 0 27] 1 1] 1 1]
           [8 [9 2 10 [13 0 54] 0 1]

@@

                [9 2 10 [6 [[0 5] 0 28] 0 119] 0 3]

@@

           [6 [3 0 13]
              [6 [5 [1 9] 0 26]
                 [6 [3 0 27]
                    [6 [3 0 54] [1 1] 1 0]

@@

           [8 [9 2 10 [13 0 55] 0 1]

@@

                [8 [8 [9 22 0 15]
                      9 2 10 [6 [0 246] 0 13] 0 2]

@@

               9 2 10 [6 [0 13] 0 5]

@@

           [6 [3 0 13]
              [6 [5 [1 10] 0 26]
                 [6 [3 0 27]
                    [6 [3 0 54]
                       [6 [3 0 108] [1 1] 1 0]

@@

           [6 [5 [1 0] 0 108]

@@

              8 [9 2 10 [13 0 55] 0 1]

@@

                [8 [9 2 10 [13 0 237] 0 3]

@@

                  [8 [8 [9 86 0 31]
                        [6 [0 1.004] [0 29] 0 13]

@@

           [6 [3 0 13]
              [6 [5 [1 11] 0 26]
                 [6 [3 0 27]
                    [6 [3 0 54]

@@

                 2 [0 28] [1 11] [0 118] [1 1] 0 5]

@@

@@

@@

   [:: +frag fragmentary subtree of noun
                    [14 [8 [9 175 0 15]
                         9 2 10 [6 0 60] 0 2]
                     8 [8 [9 342 0 15]
                        9 2 10 [6 0 60] 0 2]
       9 2 0 1]

@@

                    8 [7 [0 7] 8 [9 46 0 7]

@@
